---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# {optimizeR}

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/optimizeR)](https://CRAN.R-project.org/package=optimizeR)
[![R-CMD-check](https://github.com/loelschlaeger/optimizeR/workflows/R-CMD-check/badge.svg)](https://github.com/loelschlaeger/optimizeR/actions)
<!-- badges: end -->

This package provides a unified framework for numerical optimizers in R, particularly for inputs and outputs. 

## What is the problem? 

Look at the popular R optimizers [`nlm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/nlm.html) and [`optim()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/optim.html): The function argument in `nlm()` is called `f`, in `optim()` it is `fn`. The argument for the initial values is called `p` in `nlm()`, and `par` in `optim()`. The optimal parameters and the optimal function values in the output of `nlm()` are labeled `estimate` and `minimum`, respectively, in `optim()` it is `par` and `value`. And all is different again with [`pracma::nelder_mead()`](https://CRAN.R-project.org/package=pracma). This inconsistency is painful, especially if one wants to apply or compare different optimizers. 

## Our solution

Simply specify optimizers with `set_optimizer()` and execute them with `optimizeR()`. The outputs are in a standardized format.

For demonstration, say we want to minimize the [Ackley function](https://en.wikipedia.org/wiki/Ackley_function)...

``` r
f_ackley <- function(x) {
  stopifnot(is.numeric(x), length(x) == 2)
  -20 * exp(-0.2 * sqrt(0.5 * (x[1]^2 + x[2]^2))) -
    exp(0.5 * (cos(2 * pi * x[1]) + cos(2 * pi * x[2]))) + exp(1) + 20
}
```

... and compare three optimizers: `nlm()`, `optim()`, and `pracma::nelder_mead()`. The first two are pre-specified...

```{r optimizer1}
library("optimizeR")
opt_nlm <- set_optimizer_nlm()
opt_optim <- set_optimizer_optim()
```

... and for the latter we can use the general constructor:

```{r optimizer2}
opt_nelder_mead <- set_optimizer(
    opt = pracma::nelder_mead,
    f = "fn",
    p = "x0",
    v = "fmin",
    z = "xmin"
  )
```

Now optimize (with initial parameter vector `p = c(-1,1)`):

```{r optimization}
opt <- list(opt_nlm, opt_optim, opt_nelder_mead)
res <- lapply(opt, optimizeR, f = f_ackley, p = c(-1,1))
```

In the optimization results, `v` and `z` consistently denote the optimal function values and the optimal parameters, while optimizer-specific outputs are preserved:

```{r result}
str(res)
```

Are you surprised that the `optim()` result is different from the others? The optimizer got stuck in a local minimum. 

## Installation

The package is not yet on CRAN, but it will be soon. 

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("loelschlaeger/optimizeR")
```

